---
output: md_document
---
```{r echo=FALSE, message=FALSE }
library(tidyverse)
```

# Introduction to Spatial Analysis and GIS in R
## 2017 US EPA R User Group Workshop
### September 12, 2017
### Washington, DC

This repository contains the materials used in the Introduction to Spatial Analysis and GIS in R workshop presented at the 2017 US EPA R User Group Workshop.  

Schedule for the AM and PM workshops is below:

|Time|Topic|
|----|-----|
|8:00 AM - 8:15 AM|Welcome|
|8:15 AM - 8:30 AM|[Set-up](lessons/set_up.md)|
|8:30 AM - 9:15 AM|[Spatial data I/O](lessons/spatial_data_io.md)|
|9:15 AM - 10:15 AM|[Basic Spatial Analysis](lessons/basic_spatial_analysis.md)|
|10:15 AM - 10:30 AM | BREAK |
|10:30 AM - 11:30 AM|[Spatial Data Visualization](lessons/spatial_data_viz.md)|
|11:30 AM - 1:30 PM| LUNCH |
|1:30 PM - 1:45 PM|Welcome|
|1:45 PM - 2:00 PM|[Set-up](lessons/set_up.md)|
|2:00 PM - 2:45 PM|[Spatial data I/O](lessons/spatial_data_io.md)|
|2:45 PM - 3:45 PM|[Basic Spatial Analysis](lessons/basic_spatial_analysis.md)|
|3:45 PM - 4:00 PM| BREAK |
|4:00 PM - 5:00 PM|[Spatial Data Visualization](lessons/spatial_data_viz.md)|

#### Registrants

```{r partic, results='asis', echo=FALSE, message=FALSE}
readr::read_csv("spatial_registrants.csv")%>% 
  select(name,time) %>% 
  mutate(name = stringr::str_trim(name)) %>%
  arrange(time, name) %>%
  knitr::kable(format="markdown")
```