---
title: "05 - Explore"
author: Jeffrey W. Hollister
layout: post_page
---
Our next three lessons (Explolre, Analyze, and Visualize) don't actually split neatly into groups.  That being said, I will try my best, but there will be overlap.  For this lesson we are going to focus on some of the first things you do when you start to explore a dataset including basic summary statistics and simple visualizations with base R.

##Quick Links to Exercises and R code
- [Lesson 3 R Code](/gedr/rmd_posts/2015-01-14-05-Explore.R): All the code from this post in an R Script.
- [Exercise 1](#exercise-1): Exploring data with basic summary statistics
- [Exercise 2](#exercise-2): Using base R graphics for exploratory data analysis 

##Lesson Goals
- Be able to calculate a variety of summary statistics
- Continue building familiarity with `dplyr` and base R for summarizing groups
- Create a variety of simple exploratory plots

##Summary Statistics
There are a number of ways to get at the basic summaries of a data frame in R.  The easiest is going to be to use `summary()` which for data frames will return a summary of each column.  For numeric columns it gives quantiles, median, etc. and for factor a frequency of the terms.  Let's go back to iris and take a look.

```{r summary_data_example}
summary(iris)
```

If you want to look at the range, use `range()`, but it is looking for a vector as input.

```{r range_examp}
range(iris$Petal.Length)
```

The interquatile range can be easily grabbed with `IQR()`, again a vector is the input.

```{r iqr_examp}
IQR(iris$Sepal.Width)
```

Lastly, quantiles, at specific points, can be returned with, well, `quantile()`.

```{r quantile_example}
quantile(iris$Sepal.Length)
```

I use quantile quite a bit, as it provides a bit more flexibility because you can specify the probabilities you want to return.

```{r quantile_probs_examp}
quantile(iris$Sepal.Length, probs=(c(0.025,0.975)))
```

##Exercise 1
For this exercise, lets first spend time doing some interactive summary stats with our NLA data.

1. For this first part, explore each of the functions above (`summary()`, `range()`, `IQR()`, and `quantile()`) and get to know the `nla_data` data frame just a bit more.  Explore what ever variables you would like.

Next lets continue to add to our script.  This time we will be combining some of the past lessons with this one.  Our goal is to create a summary statistics data frame of `nla_data`.  What we want is a data frame that has min, max, mean, median, IQR, and quartiles as rows and the associated values for TURB, NTL, PTL, CHLA, and SECMEAN.  This may seem a bit daunting, but if we break it apart it gets a bit easier. This is a base R solution. 

2. Create a separate vector for each of the water quality variables.  The vector should be of length 7.
3. Combine those vectors into a data frame.

##Basic Visualization

##Exercise 2