---
title: "03 - Clean"
author: Jeffrey W. Hollister
layout: post_page
---

In my experience, most data analysis and statistics classes seem to assume that 95% of the time spent working with data is on the analysis and interpretation of that analysis and little time is spent getting data ready to analyze and, therefore, very little time is devoted to learning how to do this. However, in reality, I'd argue, the time spent is flipped with most time spent on cleaning up data and significantly less time on the analysis.    

Cleaning data (or data munging, data jujitsu, data wrangling) often happens at two points.  The first stage is getting raw data files ready to be read in by the computer.  This can be quite the challenge.  For the purposes of this workshop we are not going to focus on that side too much.  The data files we are using are in a clean enough format to be read in directly.  If you want to learn more about some best practices in maintaining good clean data files, I'd point you to two places.  First is some materials from [Data Carpentry on working with spreadsheets](https://github.com/datacarpentry/datacarpentry/blob/master/lessons/excel/ecology_spreadsheets.md) and second is [Hadley Wickham's Tidy Data paper](http://www.jstatsoft.org/v59/i10/paper).

From here on out we are assuming that we have a raw data file that is ready to be read into R and that we have already done that (we did this in [Lesson 2:Exercise 2](/introR/2015/01/14/02-Get/#exercise-2)).  While the data files are in pretty good shape at this point, there are often tasks that we need to do to select out portions of the data, rearranage the data, add new columns, merge multiple files, etc.  These are the tasks that we will focus on over the next two lessons.  This lesson will deal with using tools we have available to us in Base R.  Next lesson will focus on a new package `dplyr` that should make some of this work a bit easier.  

##Subset
But first, lets start exploring indexing to subset both observations (the rows) and variables (the columns).  

Some of the commands we know already, like `head()` and `tail()` kind of do this, but we need more control than that.  We get that with indexing.  In short indexing allows you to specify individual (or ranges) of rows and columns.  We can index data frames and vectors (and lists and matrices, but we aren't going to go into that).  Let's start with a vector example and build from there.

```{r indexing_examp}
#Create a vector
x<-c(10:19)
x
#Postive indexing returns just the value in the ith place
x[7]
#Negative indexing returns all values except the value in the ith place
x[-3]
#Ranges work too
x[8:10]
#A vector can be used to index
#Can be numeric
x[c(2,6,10)]
#Can be boolean - will repeat the pattern 
x[c(TRUE,FALSE)]
#Can even get fancy
x[x%%2==0]
```
 
So that's kinda handy.  In practice you wouldn't likely just be asking for that information to be returned to the screen, you'd probably save the output to a new object, kind of like `evens<-x[x%%2==0]`.  Let's move on now and look at indexing data frames.

Since a data frame has two dimensions, you need to specify an index for both the row and the column.  You can specify both and get a single value like `data_frame[row,column`],specify just the row and the get the whole row back like `data_frame[row,]` or get just the column with `data_frame[,column]`.  Let's work some examples.

```{r data_frame_index}
#Let's use one of the stock data frames in R, iris
head(iris)
#And grab a specific value
iris[1,1]
#A whole column
petal_len<-iris[,3]
petal_len
#A row
obs15<-iris[15,]
obs15
#Many rows
obs3to7<-iris[3:7,]
obs3to7
```

So that's the basics.  Still in the arena of kinda useful, but not really... Remember though that we can index vectors with other vectors and a data frame is just a collection of vectors (kinda) and that starts to get interesting.  Also remember that data frames have column names.  We can use those too.  Let's try it.

```{r more_data_frame_index}
#First, there are a couple of ways to use the column names
iris$Petal.Length
head(iris["Petal.Length"])
#Multiple colums
head(iris[c("Petal.Length","Species")])
#Now we can combine what we have seen to do some more complex queries
#Lets get all the data for Species with a petal length greater than 6
big_iris<-iris[iris$Petal.Length>=6,]
head(big_iris)
#Or maybe we want just the sepal widths of the virginica species
virginica_iris<-iris$Sepal.Width[iris$Species=="virginica"]
head(virginica_iris)
```

Got it?!  Probably sort of!?  I will admit that the syntax for subsetting data this way is a bit obtuse.  I did want to make sure you all saw it though becuase it comes up often in other peoples code and helps you get a better understanding, I think, of what you are doing.  That being said there are other options.  In base, `subset()` and also in `dplyr` which will be the next lesson.

```{r subset_examp}
#And redo what we did above
big_iris_subset<-subset(iris,subset=Petal.Length>=6)
head(big_iris_subset)
virginica_iris_subset<-subset(iris,subset=Species=="virginica",select=Sepal.Width)
head(virginica_iris_subset)
```

So, I think that is a big improvement in the syntax over the raw indexing and certainly lends itself to more understandable code.

##Exercise 1
This exercise is going to focus on using what we just covered on indexing and subsetting to start to clean up the National Lakes Assessment data files.  Remember to use the stickies: green when your done, red if you have a problem.

1. If it isn't already open, make sure you have the script we created, "nla_analysis.R" opened up.
2. Start a new section of code in this script by simply putting in a line or two of comments indicating what it is this set of code does.
3. Our goal for this is to create two new data frames that represent a subset of the observations as well as a subset of the data. 
4. First, from the  `nla_sites` data frame we want a new data frame that has only the following columns: SITE_ID, LON_DD, LAT_DD, STATE_NAME, WSA_ECO9, NUT_REG, NUTREG_NAME, LAKE_ORIGIN, and RT_NLA.  Name the new data frame `nla_sites_subset`.
5. Next, lets subset the water quality data from `nla_wq`.  The columns we want for this are: SITE_ID, VISIT_NO, SITE_TYPE, TURB, NTL, PTL, CHLA, and SECMEAN. Call this `nla_wq_subset`.
6. Last thing we are going to need to do is get a subset of the observations from `nla_wq_subset`.  We need only the lakes that with VISIT_NO equal to 1 and SITE_TYPE equal to "REF_Lake".  Keep the same name, `nla_wq_subset`, for this data frame.

##Merging Data
Another very common clean-up/data managment task is to be able to merge data into a single data frame.  We are going to talk about several different ways to do this.  First, is adding a new column or existing vector to a data frame with `cbind()`,`data.frame()`,or`$`. 

```{r cbind_examp}
#Create a new vector to add to the iris data frame
categories<-sample(1:3,nrow(iris),replace=T)
iris_cbind<-cbind(iris,categories)
head(iris_cbind)
#Can also use data.frame again for this
iris_cbind_df<-data.frame
head(iris_cbind_df)
#Direct assignment
iris_cbind_dollar<-iris
iris_cbind_dollar$categories<-sample(1:3,nrow(iris),replace=T)
head(iris_cbind_dollar)
```

These are all handy to know, but you need to be very careful with it.  First, the data in the vector you are adding needs to be in the same order as the observations in the data frame.  Also it needs to have the same number of rows (or divisible by that number of rows).  

The next thing we might want to do is add some new rows to a data frame.  This is very handy as you might have data collected and entered at one time, and then additional observations made later that need to be added.  So with `rbind()` we can stack two data frames with the same columns to store more observations.  

```{r rbind_examp}
#Let's first create a new small example data.frame
rbind_df<-data.frame(a=1:3,b=c("a","b","c"),c=c(T,T,F),d=rnorm(3))
#Now an example df to add
rbind_df2<-data.frame(a=10:12,b=c("x","y","z"),c=c(F,F,F),d=rnorm(3))
rbind_df<-rbind(rbind_df, rbind_df2)
rbind_df
```

Now something to think about.  Could you add a vector as a new row?  Why/Why not? When/When not?

Let's go back to the columns now.  What we can currently do is add columns that need to be same length, or length that is divisible by number of rows in the original, and need to be in the same order.  But it is very common to have to datasets that are in different orders and have differing numbers of rows.  What we want to do in that case is going to be more of a database type function and join two tables based on a common column.  A common way to do that in base R is with `merge()`.   So let's contrive another example by creating a dataset to merge to `rbind_df` that we created above.

```{r merge_example}
# Contrived data frame
rbind_df_merge_me<-data.frame(a=c(1,3,10,11,14,6,23),x=rnorm(7),names=c("bob","joe","sue",NA,NA,"jeff",NA))
# Create merge of matches
rbind_df_merge_match<-merge(rbind_df,rbind_df_merge_me,by="a")
rbind_df_merge_match
# Create merge of matches and all of the first data frame
rbind_df_merge_allx<-merge(rbind_df,rbind_df_merge_me,by="a",all.x=TRUE)
rbind_df_merge_allx
```

##Exercise 2
In this exercise we are going to practice merging our NLA data.  You will remember that we have two datasets, one on water quality and one on site information.  We selected some info out of the water quality data that we didn't select out of the site info so we have two data frames, with differing numbers of rows and unknown order.  Use your stickies!

1. This is the only task we have for this exercise.  Add to your script a line (or more if you need it) to create a new data frame, `nla_data`, that is a merge of `nla_site_subset` and `nla_wq_subset`, but with all lines in `nla_wq_subset` preserved in the output.
2. If that goes quickly, feel free to explore `rbind()` some.

##Reshape and Modifying
So we now have a data frame that is set up pretty well to facilitate some future analysis.  That being said, different analyses and visualizations are probably going to require some additional monekying around with our data.  This requires us to reshape, modify, aggregate, or split up our dataset.  There are many ways to do this and a lot depends on the format of the original data.  We will focus on a couple: transposing and aggregation.  There is a lot more you can do.  Again, the best source for packages and examples on this is Hadley Wickham.  In particular I would look at the `reshape` package and the associated links:

- [`rehsape` pakcage](http://had.co.nz/reshape/)
- [JSS paper](http://www.jstatsoft.org/v21/i12/paper)

We will explore another one of Hadley's packages, `dplyr` soon, but right now let's look at some basic reshaping in base R.  We will show how to transpose and aggregate.  We already know how to do basic subsetting.  These techniques will add to our toolbox.

To transpose a data frame all we need is the `t()` function. 


```{r transpose_example}
#Look at are contrived example again
rbind_df_merge_match
#And the transpose of this
t(rbind_df_merge_match)
```

I find that I don't use transpose a lot, but when I do it is usually in combination with other summary and aggregation methods.  It can also be useful with data in a long form where each row does not represent a single observation and the observation id is duplicated.  Getting that to work with base R takes a lot of work.  We will talk a bit more about that in the next lesson.

Let's now look at some aggregation commands.  We will focus on the wrapper functions `aggregate()` and `by()`.  These provide a more convenient way to access the `apply` family of functions.  I will introduce those, but not go into too much detail.  

There are many apply functions and they include: `apply()`, `lapply()`, `sapply()`, `mapply()`, `tapply()`,  and `eapply()` (I think that is all of them...).  Since we are working with data frames almost exclusively in this workshop, the two that I seem to use the most with data frames are `apply()` and `tapply()`.  `apply()` allows you to run a function by row or by columns.  For instance, you could have multiple columns that you wanted get an average across, weekly temperature, perhaps:

```{r apply_examp}
temp_df<-data.frame(id=1:100,week1=runif(100,20,25), week2=runif(100,19,24), 
                    week3=runif(100,18,26), week4=runif(100,17,23))
head(temp_df)
temp_df$month_mean_temp <- apply(temp_df,1,mean)
head(temp_df)
```

```{r tapply_examp}
```

```{r aggregate_examp}
```


##Exercise 3


